Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-25 Report/Sections/discussion.tex
Commit before restructure!

In particular, we use $Z^{*(1)}$ and $Z^{*(2)}$ based on the DFT as defined previously. These summary statistics prove to be vital when decoding the behavior of a killer whale. This is evidenced by both our simulation study, where the sub-dive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.

@@

An example of balancing model complexity with efficient fitting is presented in our simulation study. Sub-dive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train was the CarHMM-DFT, which had no hierarchical component. Even still, the CarHMM-DFT had near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale.
However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.

@@

%We used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to simultaneously distinguish three distinct sub-dive behaviors and two dive types. The DFT component proved useful in determining the sub-dive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each sub-dive state was separated by an order of magnitude. Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data. The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and sub-dive behaviors was selected in an ad-hoc manner. There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.

>>>>

In particular, we use $Z^{*(1)}$ and $Z^{*(2)}$ based on the DFT as defined previously. These summary statistics prove to be vital when decoding the behavior of a killer whale. This is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.

@@

An example of balancing model complexity with efficient fitting is presented in our simulation study. subdive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train was the CarHMM-DFT, which had no hierarchical component. Even still, the CarHMM-DFT had near-perfect accuracy when decoding subdive behavioral states of the simulated whale.
However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.

@@

%We used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to simultaneously distinguish three distinct subdive behaviors and two dive types. The DFT component proved useful in determining the subdive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each subdive state was separated by an order of magnitude. Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and subdive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data. The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and subdive behaviors was selected in an ad-hoc manner. There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.

### 2020-07-25 Report/Sections/intro.tex
Commit before restructure!

Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
Some techniques, such as time series, are highly parametric. 
While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
At the other extreme are black box type methods of machine learning.  
While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.  
The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).

Our goal is to develop movement analysis methods that are flexible but also use subject area knowledge to provide insight into curve data with complex structure.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure via Markov processes.  
While our methodology is generally applicable, [rephrase - awkward] we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
Figure \ref{fig:data} shows the data we will analyze.
The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions, as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

@@

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.  Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

@@

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017}.
One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. 
%
%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioral process.
%However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
%while maintaining the structure needed to run the forward algorithm. 
%It also only adds one additional parameter per possible hidden state.

>>>>

%Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
%Some techniques, such as time series, are highly parametric. 
%While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
%At the other extreme are black box type methods of machine learning.  
%While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
%Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.  
%The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).

Often, movement data can be viewed as a set of many curves.   In Morris, each child in the data set is associated with two curves from two different days.   In our data, we will consider curves from a sequence of dives of one killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada. 
These types of data are amenable to techniques of Functional Data Analysis (FDA).
FDA was originally developed for curves that can be considered independent replicates  (Ramsay ref, a theoretical ref).  
However, the structure of the data is often more complicated.
For instance, in Morris, clearly, the two curves from the same child are dependent.
And in our dive data, the sequence of days exhibits temporal dependence.

More complicated structures in FDA have been considered in the form of hierarchical ...  typically incorporating random effects.
Time series of curves  ...  RELATED would be appropriate here.
Here, we consider Markov modelling of a curve sequence.  Common in animal movement because it gives biological insight.

Our goal is to develop movement analysis methods that are flexible but also use subject area knowledge to provide insight into curve data with complex structure.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure via Markov processes.  
While our methodology is generally applicable, we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
Figure \ref{fig:data} shows the data we will analyze.
The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions, as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

@@

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.  Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

@@

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017,Mcclintock:2020}.
One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. 
%
%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioral process.
%However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
%while maintaining the structure needed to run the forward algorithm. 
%It also only adds one additional parameter per possible hidden state.

### 2020-07-25 Report/Sections/results.tex
Updates from Overleaf
Commit before restructure!

The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada. Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale, as well as the distribution of accelerometer data within each behavioral state, is important to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

@@

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time periods were ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed dives within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

@@

We used a hierarchical HHM, where the coarse-scale observations were a time series of dive durations in seconds, and the fine-scale observations were the within-dive acceleration data (Figure \ref{fig:CarHHMM-DFT}). The coarse-scale level is comprised of an HMM with hidden states corresponding to dive types. The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

The fine-scale level model subdive behavior and is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window). The acceleration exhibits significant sinusoidal behavior, thus the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window. Unlike the simulation study, where we had one time series of acceleration data, we have here acceleration data in three dimensions. These three dimensions represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags (Citation?). The following two sets of observations $z^*$ were calculated for modeling:
%
$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar. We calculate $z_t^{*(2)}$ by summing the first 10 Fourier modes, and corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.

There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (see the supplementary material for a lag plot), thus auto-correlation was directly modeled into the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$ using a CarHMM. We assumed the error to be Normally distributed with the following parameters:
%
$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)}$$
$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right]$$
%
where $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.

While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is less clear. Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$. In particular, the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean and variance:
%
$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the sub-dive state $X_{t,t^*}$.

Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.

@@

The estimates for the emission distribution parameters suggest that this whale has two distinct dive behaviors (Table \ref{table:emis_dists},  figure \ref{fig:coarse_emis}). Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to such short dives, including resting (CITATION, you can ask Ian and Sarah for help here - look at the Tennessen paper with HMM). Dive type 2 is longer and deeper. These types of deep sustained dives have been associated with foraging behaviors in killer whales (CItation Tennessen maybe HMM and the other Tennesen with foraging data).

The three subdive behaviors also appeared distinct, particularly with respect to the fourier sum (Table \ref{table:emis_dists},  Figure  \ref{fig:fine_emis}). Sub-dive type 1 corresponds to gliding and less overall activity compared to the other behavioral states. The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavioral state 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state. Sub-dive state 3, on the other hand, corresponds to vigorous swimming activity, as the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ for every component is much higher than every other state. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds. Finally, sub-dive state 2 corresponds to a moderate amount of activity, as almost every parameter estimate is between the other two behavioral states.

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.849  &  0.151 \\
0.907  &  0.093
\end{pmatrix},$$
$$\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix},$$
%
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.724 & 0.276 & 0.000 \\
0.057 & 0.887 & 0.056 \\
0.000 & 0.247 & 0.753
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.871 & 0.129 & 0.000 \\
0.135 & 0.829 & 0.036 \\
0.000 & 0.246 & 0.754
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix}, \qquad
\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}.$$
%
The coarse-scale stationary distribution indicates that most dives are short type 1 dives. The coarse-scale transition probability matrix further indicates that the whale makes multiple short type 1 dives before doing only one (or a few) long type 2 dive. The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active sub-dive type 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}. Such strategy could explain why sub-dive state 1 makes up 14\% of shallow dives of our killer whale, but it makes up 48\% of deep dives. Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 5 selected dives. The supplementary material also shows the estimated \textit{probability} of each dive time and sub-dive hidden state.

@@

We also plotted separate histograms of each feature where each observation was weighted by the probability that the whale was in a particular hidden state. This empirical distribution was then plotted over the fitted probability distribution function of that feature and hidden state. If the model is correct, then the histogram of each feature and hidden state should closely resemble its corresponding fitted probability distribution. See Figure \ref{fig:empirical_dist} for an example of the dive duration. Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study. See the supplementary material for empirical distributions of every feature and every hidden state.

>>>>

In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada. Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, which can be helpful for conservation efforts. However, studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

@@

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

@@

We used a hierarchical HHM, where the coarse-scale observations were a time series of dive durations in seconds, and the fine-scale observations were the within-dive acceleration data (Figure \ref{fig:CarHHMM-DFT}). The coarse-scale level was comprised of an HMM with hidden states corresponding to dive types. The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

The fine-scale acceleration data exhibits significant sinusoidal behavior, thus the fine-scale observations $Y^*$ were transformed using the DFT summary statistics of a two-second sliding window. Unlike the simulation study, where the time series of acceration was one-dimensional, the acceleration data here was collected in three dimensions. These three dimensions represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. The following two sets of observations $z^*$ were calculated:
%
$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar. We calculated $z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.

There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (section S-2.1), thus auto-correlation was directly modeled into the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$ using a CarHMM. We assumed $\mathbf{Z}^{*(1)}_{t,t^*}$ to be Normally distributed with the following parameters:
%
$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)}$$
$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right]$$
%
where $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.

While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is less clear. Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$. In particular, the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean and variance:
%
$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the subdive state $X_{t,t^*}$.

Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ subdive behaviors heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three subdive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.

@@

The estimates for the emission distribution parameters suggest that this whale has at least two distinct dive behaviors (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to such short dives, including resting (CITATION, you can ask Ian and Sarah for help here - look at the Tennessen paper with HMM). Dive type 2 is longer and deeper. These types of deep sustained dives have been associated with foraging behaviors in killer whales (Citation Tennessen maybe HMM and the other Tennesen with foraging data).

The three subdive behaviors also appeared distinct, particularly with respect to their Fourier sums, as the mean of $Z^{*(2)}$ is separated by an order of magnitude between the three subdive types (Table \ref{table:emis_dists_CarHHMM-DFT},  Figure  \ref{fig:fine_emis}). Subdive type 1 corresponds to gliding and less overall activity compared to the other subdive types. The mean of $Z^{*(2)}$ and the variance of $\mathbf{Z}^{*(1)}$ in this state is smaller than the other subdive types. The auto-correlation of $\mathbf{Z}^{*(1)}$ also is higher for subdive type 1 than every other subdive type.
Subdive type 2 is characterized by a mean Fourier sum ($Z^{*(2)}$) one order of magnitude high than subdive type 1. Acceleration has about twice the variance in this subdive type compared to subdive type 2, and slightly lower auto-correlation. Subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behavior in acceleration is characteristic of this behavior marine mammals \citep{Simon:2012}.
Finally, subdive type 3 corresponds to vigorous swimming activity, especially as the killer whale submerges or surfaces. The mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ is much higher in subdive type 3 compared to every other state. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this subdive type, implying more variation in acceleration.

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix},$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix},$$
%
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \qquad
\hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}.$$
%
The coarse-scale stationary distribution indicates that most dives are short type 1 dives. The coarse-scale transition probability matrix further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive. The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active subdive type 1 when performing long deep dives than when performing short shallow dives. Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}. Such strategy could explain why subdive type 1 makes up 9\% of shallow dives for this killer whale, but it makes up 40\% of deep dives. Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 6 selected dives. The supplementary material also shows the estimated \textit{probability} of each dive time and subdive hidden state (section S-2.2).

@@

We also plotted histograms of each feature where observations were weighted by the probability of a particular hidden state. This empirical distribution was then plotted over its corresponding fitted emission distribution (Figure \ref{fig:empirical_dist}). Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study. See the supplementary material for empirical distributions of every feature and every hidden state.
